\chapter{What Business Is For}\label{ch:value-creation}

% Scope: value, economic profit as a signal, ROIC vs WACC as the master test.
% Cross-link: LTV:CAC is the per-customer expression of ROIC > WACC (\cref{ch:customer-economics}).

Every financial metric in the book --- LTV:CAC, ROAS, NOPAT, terminal value --- is a
proxy for one underlying question: does this activity generate returns above the cost of the
capital it consumes? Economic profit answers that question directly; the rest of the book
traces how it surfaces in each functional domain.

\section{Economic Profit}\label{sec:economic-profit}
% complexity: 3

Did we earn more than a fair return on the capital we deployed? Accounting profit ignores the
equity charge; economic profit does not. This gap is where most businesses quietly destroy value.
Lenders charge interest, and that charge appears in the income statement. Equity holders also
have an opportunity cost --- they could invest in the market instead. Accounting profit omits
that charge. Economic profit subtracts it: a positive number means the business earned above its
all-in cost of capital; a negative number means it would have been better to return the money to
investors, even if accounting profit was positive.

Let \(\mathrm{NOPAT}\) be net operating profit after tax (operating income minus the tax shield on
operations, before financing), \(\mathrm{WACC}\) the weighted-average cost of capital, and
\(\mathrm{IC}\) the invested capital (equity plus net debt deployed in the business). Then:
\begin{equation}\label{eq:economic-profit}
  \mathrm{EP}
    \;=\; \mathrm{NOPAT} - \mathrm{WACC}\times\mathrm{IC}
    \;=\; (\mathrm{ROIC}-\mathrm{WACC})\times\mathrm{IC}.
\end{equation}
The right-hand form separates \emph{spread} (ROIC minus WACC) from \emph{scale} (IC): a high
spread on a small capital base earns less EP than a moderate spread on a large one. Both levers matter.

The formula holds only when NOPAT is clean: operating items only, after-tax, stripped of
one-time charges. WACC must match the riskiness of the cash flows being evaluated --- using a
corporate WACC for a speculative new division understates the hurdle. Invested capital must
capture all capital consumed, including operating leases and capitalised R\&D where material.
Goodwill from acquisitions is especially dangerous: including it inflates IC and depresses ROIC
post-deal, making an accretive deal look value-destroying.

\begin{example}
The running SaaS business earns \(\mathrm{NOPAT}\approx\money{600000}\) on invested capital
of \(\money{3000000}\), giving \(\mathrm{ROIC}=20\%\). With \(\mathrm{WACC}=11\%\):
\[
  \mathrm{EP}
    = (\underbrace{0.20}_{\text{return earned}} - \underbrace{0.11}_{\text{return required}})
      \times \money{3000000}
    = \money{270000}.
\]
The business is creating \money{270000} of value per year --- the spread of \qty{9}{\percent}
on the capital base. Doubling IC while holding ROIC constant doubles EP; maintaining ROIC while
WACC rises (e.g.\ because beta increases) erodes it without any operational change.
\end{example}

\begin{admonition}{Warning}
Free cash flow and economic profit are not the same thing. FCF can be positive while EP is
negative: a mature business harvesting a declining franchise generates cash but may earn below its
cost of capital. Conversely, a high-growth company with strongly negative FCF may have strongly
positive EP if it is deploying capital at returns well above WACC. The two metrics answer
different questions; neither substitutes for the other.
\end{admonition}

Stern Stewart's EVA (Economic Value Added) is a branded implementation of \cref{eq:economic-profit}
with adjustments to reported accounting figures (capitalising R\&D, reversing LIFO reserves, etc.)
to produce an economic capital number. MVA (Market Value Added) cumulates the present value of
expected future EP --- it is equivalent to the premium of market capitalisation over invested
capital. The DCF--EP equivalence is exact: the NPV of all future EPs, discounted at WACC, equals
the NPV of all future free cash flows \parencite{brealey2020}. Use whichever is easier to
estimate; they agree when inputs are consistent.

The ROIC--WACC spread of \cref{eq:economic-profit} recurs throughout the book: at the customer level
as LTV:CAC (\cref{ch:customer-economics}), at the firm level as the advantage spread
(\cref{ch:advantage-and-disruption}), and at the portfolio level in scenario-weighted DCF
(\cref{ch:valuation}). \textcite{ross2022} and \textcite{brealey2020} develop the corporate-finance
foundations; \textcite{berk6thcorporate} bridges to market-based valuation.

\section{ROIC vs.\ WACC --- the Master Test}\label{sec:roic-wacc}
% complexity: 3

Is the business compounding value or consuming it? One ratio --- ROIC relative to WACC ---
answers that question for the whole enterprise and for any proposed investment within it.
Capital is not free. When a business earns more on each dollar deployed than investors require to
supply that dollar, it compounds wealth. When it earns less, it destroys wealth even if it is
profitable in the accounting sense. The distance between ROIC and WACC, multiplied by the capital
base, is economic profit (\cref{eq:economic-profit}).

Return on invested capital is simply the ratio of operating return to capital deployed:
\begin{equation}\label{eq:roic}
  \mathrm{ROIC} \;=\; \frac{\mathrm{NOPAT}}{\mathrm{IC}}.
\end{equation}
Value is created when \(\mathrm{ROIC}>\mathrm{WACC}\), destroyed when \(\mathrm{ROIC}<\mathrm{WACC}\),
and neutral when equal. Growth amplifies whichever condition holds: investing more capital at
\(\mathrm{ROIC}>\mathrm{WACC}\) creates proportionally more EP; investing at
\(\mathrm{ROIC}<\mathrm{WACC}\) destroys it faster.

ROIC is sensitive to the denominator. A capital-light business (high asset turnover) can show
high ROIC with modest margins; a capital-heavy one must earn thick margins just to clear WACC.
Both readings are legitimate --- what matters is the spread, not the absolute level of either
metric. ROIC also fluctuates with the business cycle; a single year's figure can mislead; use
a normalised or through-cycle average.

\begin{example}
Continuing the running example: \(\mathrm{ROIC}=\money{600000}/\money{3000000}=20\%\) against
\(\mathrm{WACC}=11\%\). The spread is \qty{9}{\percent} --- strongly value-creating.

At the customer level, \(\mathrm{LTV}:\mathrm{CAC}=\money{3150}/\money{600}=5.25\times\)
is the per-customer form of the same test: lifetime return on customer-acquisition capital,
compared implicitly against the hurdle \(\mathrm{LTV}:\mathrm{CAC}=1\). The mechanics of that
calculation are in \cref{ch:customer-economics}; the equivalence is that both measure ROIC on a
defined capital deployment over its full life.
\end{example}

\begin{admonition}{Warning}
Optimising ROIC by shrinking the denominator is a reliable path to destroying the business.
Cutting invested capital --- postponing maintenance, starving marketing, running down working
capital --- raises measured ROIC in the short run while depleting the productive base that
generates future NOPAT. The metric then looks better precisely as the business deteriorates.
A rising ROIC with flat or falling NOPAT is a warning sign, not a success.
\end{admonition}

EVA and MVA (introduced in \cref{sec:economic-profit}) quantify the cumulated effect of
sustained ROIC--WACC spreads. In a DCF framework, a company that sustains \(\mathrm{ROIC}>\mathrm{WACC}\)
indefinitely has a terminal value above book value; one that earns exactly WACC has a terminal
value equal to book. The McKinsey valuation framework makes this the central organising
principle \parencite{damodaran2012}. At the acquisition level, a deal creates value only if the
post-deal ROIC on total invested capital (including the premium paid) exceeds WACC --- the
rationale for rigorous integration synergy modelling before signing.

ROIC anchors the corporate-finance treatment in \cref{ch:corporate-finance-core} and the full DCF
build in \cref{ch:valuation}. The cost of capital WACC is derived from first principles in
\cref{ch:risk-and-capital-structure}. \textcite{damodaran2012} provides the standard treatment of
ROIC-based value drivers.

\section{Stakeholder vs.\ Shareholder Value}\label{sec:stakeholder-value}
% complexity: 3

Whose interests should the business optimise for? The answer shapes capital allocation, executive
incentives, and strategy --- and it is genuinely contested. The shareholder-primacy view holds
that management's fiduciary duty runs to equity owners; any stakeholder benefit is legitimate
only insofar as it serves shareholder returns. The stakeholder view holds that customers,
employees, suppliers, and communities are co-creators of value whose interests constrain and
ultimately enable those returns. The two positions agree on more than the debate suggests ---
sustainable shareholder value requires satisfied customers and motivated employees --- but they
diverge sharply on governance, pay structures, and acceptable trade-offs.

The shareholder-primacy position is stated precisely in \textcite{ross2022}: maximise the present
value of future free cash flows distributable to equity. The stakeholder rebuttal --- Freeman's
strategic management framework --- holds that stakeholder relationships are themselves
value-generating assets; ignoring them understates the long-run cash flows:
\begin{description}
  \item[Shareholder primacy] Equity holders are residual claimants; maximising their wealth is
    the unambiguous objective. All other stakeholder relationships are inputs, priced by market
    competition. \parencite{brealey2020}
  \item[Stakeholder theory] Strategy must account for the claims of customers, employees,
    suppliers, communities, and regulators. Ignoring these relationships generates hidden
    liabilities that materialise as regulatory risk, talent attrition, and customer defection.
  \item[Service-dominant logic] \textcite{vargo2004service} reframe value as co-created between
    the firm and its customers in the act of service: the product is an operant resource, not a
    finished good. Under this view, economic profit accrues from relationships, not transactions.
\end{description}

Shareholder primacy breaks down empirically when externalities are large (e.g.\ pollution), when
relationships are the primary asset (professional services, platforms), or when short-term
earnings pressure destroys the long-term investment capacity that equity value depends on. The
stakeholder model breaks down when the constituency list is so broad that management has no
clear optimisation target and can rationalise any decision as serving someone.

\begin{example}
Applied to the running SaaS business: shareholder-primacy reasoning supports accepting every
customer-acquisition dollar whose expected LTV:CAC clears the WACC hurdle (\cref{eq:roic}).
Stakeholder reasoning adds: does the product actually improve the customer's outcome, and is
the contribution-margin structure fair to the employees who deliver it? In practice, the SaaS
model aligns both: subscription renewal is voluntary, so sustained LTV requires genuine value
delivery. Misalignment typically appears in enterprise lock-in contracts or opaque pricing ---
not in transparent usage-based SaaS.
\end{example}

\begin{admonition}{Warning}
Treating ESG reporting as a substitute for economic-profit analysis conflates two distinct
questions. A high ESG score does not imply \(\mathrm{ROIC}>\mathrm{WACC}\); a low one does not
imply the opposite. Each deserves its own rigorous metric: economic profit for capital
productivity, explicit social-impact measurement for externalities.
\end{admonition}

The Friedman doctrine --- ``the social responsibility of business is to increase its profits''
(\emph{New York Times}, 1970) --- remains the strict shareholder-primacy anchor. The 2019
Business Roundtable statement, signed by 181 CEOs, explicitly rejected shareholder primacy in
favour of stakeholder commitments. Whether this represents a genuine shift in governance or
reputational signalling is actively debated; the empirical literature finds mixed evidence on
whether stakeholder commitments improve or impair shareholder returns over long horizons
\parencite{kotler2021}.

Service-dominant logic (\cref{sec:stakeholder-value}) and the jobs-to-be-done framework
(\cref{ch:segmentation-targeting-positioning}) converge on a common implication: the unit of
value is not the product but the outcome the customer achieves. Firms that design around outcomes
rather than features tend to earn higher retention and lower CAC --- which shows up as higher
LTV:CAC, which is economic profit at the customer level. The frameworks are complementary, not
competing, when the time horizon is long enough.

The tension between shareholder and stakeholder value recurs in governance (\cref{ch:organizational-behavior})
and in the pricing ethics discussion of \cref{ch:pricing}. \textcite{vargo2004service} provide the
service-dominant logic foundation; \textcite{kotler2021} survey the marketing-strategy implications.

\section{ESG as a Capital-Allocation Discipline}\label{sec:esg-capital}
% complexity: 3

When does an environmental, social, or governance factor belong inside a capital-allocation
decision rather than a separate sustainability report? The discipline is rigorous: an ESG factor
earns a place in the capital-allocation calculus only if it measurably affects NOPAT, WACC, or
invested capital --- the three levers of \cref{eq:roic} and \cref{eq:economic-profit}. Factors
that do not meet that test may be worth disclosing, but they belong to a different analysis.
The ISSB's IFRS S1 and S2 standards formalise this financial-materiality lens at the reporting
level: S1 requires disclosure of sustainability risks that affect enterprise value; S2 extends
that specifically to climate. The European double-materiality standard (ESRS under the CSRD)
goes further, requiring disclosure when the impact on society is material \emph{even if} no
financial effect is yet detectable. For capital-allocation purposes, the financially material
subset is the operative one.

In practice, ESG factors enter the capital model through three channels. First, a risk reprices
the cost of capital: a company with material transition risk (regulatory carbon costs) or
physical climate exposure will find lenders and equity investors demanding a higher risk premium,
raising WACC directly. Second, an operational inefficiency or liability suppresses NOPAT: health
and safety failures, supply-chain disruption from climate events, and discrimination suits all
reduce after-tax operating profit. Third, a factor inflates invested capital without a
proportional return: stranded assets, environmental remediation obligations, and deferred
infrastructure spending all appear in IC once accounting is cleaned up. When any of these
channels is non-trivial, the ESG analysis is inseparable from the economic-profit analysis.

\begin{example}
The running SaaS business faces one ESG factor that clearly meets the financial-materiality
threshold: a gender pay gap concentrated in the engineering team. The mechanism is via NOPAT ---
underpaid engineers leave, triggering replacement searches. Each mid-level engineering
departure costs an estimated \money{80000} in recruiting, onboarding, and productivity loss.
A pay-equity audit and remediation programme costs \money{20000} and prevents one departure per
year.

The decision is a standard NPV application (\cref{eq:npv}). The annual benefit is the avoided
replacement cost of \money{80000}; the annual additional cost is the audit of \money{20000}; the
net annual cash inflow is \money{60000}. Treating this as a perpetuity at the firm's \qty{11}{\percent}
WACC,
\[
  \mathrm{NPV} \;=\; \frac{\money{80000} - \money{20000}}{0.11} \;\approx\; \money{545000}.
\]
The investment is strongly NPV-positive --- not because of ESG optics, but because the
financial-materiality test is satisfied. Conversely, a carbon-offset programme that costs
\money{30000} per year and has no measurable effect on NOPAT, WACC, or IC fails the test and
belongs in the CSR budget, not the capital plan.
\end{example}

\begin{admonition}{Note}
The Warning in this chapter that a high ESG score does not imply \(\mathrm{ROIC}>\mathrm{WACC}\)
--- and vice versa --- applies here directly. The pay-equity example above passes the financial-
materiality test; most reported ESG scores are aggregates of factors, many of which do not.
The discipline is to disaggregate: score each ESG factor against the three capital-model channels
(NOPAT, WACC, IC) before including it in any projection.
\end{admonition}

The stakeholder relationships that \textcite{freeman2010strategic} identifies as strategic assets
are precisely the ones that pass the financial-materiality test: employees whose engagement
sustains NOPAT, customers whose trust sustains retention and limits CAC, and regulators whose
goodwill avoids licence risk that would inflate WACC. \textcite{vargo2004service}'s co-creation
lens adds that the employee relationship is an operant resource --- it cannot be replaced by a
transaction; its deterioration reduces the productive capacity that NOPAT measures. Both
frameworks, read through the capital-allocation lens, arrive at the same conclusion: ESG is not
a parallel reporting exercise but an upstream input to the economic-profit calculation when
materiality is satisfied.

The ESG data infrastructure required to run this analysis at scale is a chapter-17 problem
(\cref{ch:data-and-ai-strategy}): which sustainability metrics are decision-grade, how they are
collected, and how double-materiality assessments are operationalised as a data workflow. Board
oversight of what gets disclosed and what enters the capital plan is addressed in
\cref{ch:organizational-behavior}.
